Redundant calibration as a complex optimization problem

### 2017-01-24 paper.tex
added new layout figures

\begin{abstract}
In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. 
We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework.
We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton.
Traditionally redundant calibration is implemented via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration.
The Gauss-Newton algorithm is, however, still the de facto standard that is currently being used to perform redundant calibration.
We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating 
direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. 
We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method 
and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

Two ways of speeding up redundant calibration have recently been proposed. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities \citep{Wijnholds2012,Marthi2014}. Note that the fact that \citet{Marthi2014} make use of the ADI method is not immediately clear
from the paper itself. We will elaborate on this in Section~\ref{sec:adi}. The second is to use the preconditioned conjugate gradient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem 
to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before.

In this paper we propose to use the complex framework introduced by \citet{Smirnov2015} instead of the traditional amplitude-and-phase approach used by \textsc{lincal}.
We also propose to use the Levenberg-Marquardt (LM) gradient-based algorithm instead of the Gauss-Newton (GN) algorithm as it exhibits better convergence properties.
Using the LM algorithm reduces the number of algorithmic layers that are needed to perform redundant calibration. 
We also compare two recent redundant calibration speedup algorithms with each other, namely the ADI \citep{Marthi2014} and PCG \citep{Liu2010} methods. In this paper, we decided to focus on the complex framework proposed by \citet{Smirnov2015} rather than the traditional phase-and-amplitude approach as it 
allows us to trivially prove that the ADI method proposed by \citep{Marthi2014} is merely a special case of the more general LM algorithm.

@@

We start the paper by presenting some useful preliminaries in Section~\ref{sec:preliminaries}. In Section~\ref{sec:p_statement}, we formulate redundant calibration 
as a least squares problem, i.e. Eq~\eqref{eq:least_squares_red}. In Section~\ref{sec:main_sec} we use the complex framework proposed by \citet{Smirnov2015} to derive a numerical
algorithm (see Eq.~\eqref{eq:GN_update} to Eq.~\eqref{eq:update}), which we can use to solve Eq~\eqref{eq:least_squares_red}. In Section~\ref{sec:results}, we compare the computational complexity of two recently proposed redundant calibration speedup algorithms with one another.
We end the paper with some final conclusions.

@@

\subsubsection{Preconditioning}
\label{sec:precon}
Preconditioning is a technique which allows us to improve the condition number of a matrix. Let $\bM$ be a positive-definite Hermitian matrix.
We can now indirectly solve for $\bx$ by using 
\begin{equation}
\label{eq:prec}
\bM^{-1}\bA\bx = \bM^{-1}\bb.
\end{equation}
If $\bM$ is a good preconditioner then $\kappa(\bM^{-1}\bA) \ll \kappa(\bA)$, i.e. a good preconditioner lowers the condition number of a matrix which in turn increases the convergence speed 
of CG (see Eq.~\eqref{eq:cg_bound}). In this paper, we denote matrix inversion with $(*)^{-1}$. The best possible choice of $\bM$ would therefore be $\bA$, which would lead to $\kappa$ becoming unity (as the largest and smallest eigenvalue of the Identity matrix are equal to one).
If $\kappa$ is one then Eq.~\eqref{eq:cg_bound} simplifies and becomes $\mathbb{O}(m)$. However, if we knew the inverse of $\bA$ then obtaining $\bx$ would be trivial and there would be no need for the CG method in the first place. 
In short, a good preconditioner $\bM$ must approximate $\bA$ fairly accurately. However, if it is not much less expensive to 
compute the inverse of $\bM$ than that of $\bA$ we actually gain nothing (see Eq.~\eqref{eq:prec}).

@@

\subsection{Least Squares}
\label{sec:ls}
The method of nonlinear least squares is tanatamount to solving the following optimization problem:  
\begin{equation}
\label{eq:least_squares}
\min_{\bzeta} \Lambda(\bzeta) = \min_{\bzeta} \|\brho\|_F^2 = \min_{\bzeta} \|\bdelta - \bnu(\bzeta)\|_F^2, 
\end{equation}
where $\Lambda$ is known as the objective function, $\brho$ is the residual vector, $\bdelta$ is the observed data vector, $\bnu$ is the model-predicted data vector and $\bzeta$ is the model-parameter vector that minimizes Eq.~\eqref{eq:least_squares}.
In Eq.~\eqref{eq:least_squares}, we assume that the optimization problem is real valued.
The main objective of least squares is to find the model-parameters which minimize the squared difference between the observed data and the model-predicted data.

@@

The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a standard deviation $\sigma$, where   
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}}. 
\end{equation}
Here $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

Recall that redundant baselines sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. 
Let us now group all the baselines that measure the same $uv$-modes in a redundant array into the same redundant baseline group. Assigning the baselines of a redundant array to different redundant baseline groups has the advantage that it reduces the number of single index values we require. We now only require single index values for each redundant baseline group instead of a unique single index label for 
each baseline. Let $\phi_{pq}$ be the function that maps composite antenna pairs to the appropriate single redundant baseline group index. Note that $\phi_{pq}$ is not symmetric as 
$\phi_{pq} = 0$ if $p>q$. We can also define the following symmetric variant of $\phi_{pq}$:
\begin{equation}
\zeta_{pq} = 
\begin{cases}
\phi_{pq}~\textrm{if}~p \leq q\\
\phi_{qp}~\textrm{if}~p>q
\end{cases}.
\end{equation} 
It is possible to construct a simple analytic expression for $\zeta_{pq}$ when our array is in an east-west regular configuration, namely $\zeta_{pq} = |q-p|$. 
It becomes increasingly difficult to construct analytic expressions of $\zeta_{pq}$ for more complicated array layouts. 
We will refer to $\zeta_{pq}$ as the symmetric geometric function. The empirically constructed symmetric geometry
functions of three different redundant layouts are depicted in Fig.~\ref{fig:geometry_function}. We denote the range of $\zeta_{pq}$ with $\mathcal{R}(\zeta_{pq})$. The maximal element 
that $\zeta_{pq}$ can ascertain is denoted by $L$ and can be interpreted as the maximal number of unique redundant baseline groups which can be formed for a given 
array layout. Alternatively, we can interpret the symmetric geometry functions in Fig.~\ref{fig:geometry_function} as geometry matrices. In this alternative paradigm, $\zeta_{pq}$ denotes a matrix entry. The dimension of these geometry 
matrices are $N\times N$.

@@

\subfigure[Hexagonal: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./HEX_phi.pdf}\label{fig:HEX_phi}}
\subfigure[Square: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./SQR_phi.pdf}\label{fig:SQR_phi}}
\subfigure[Regular east-west: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./REG_phi.pdf}\label{fig:REG_phi}}
\caption{In the top panel we have three different redundant antenna layouts, namely a hexagonal (left), square (middle) and a regular east-west (right) layout. 
In the bottom panel we have the respective symmetric redundancy geometry functions for the layouts on the top panel. We used 91 antennas
to construct the the hexagonal layout, while a 100 antennas where used in the square and east-west layouts. The maximal amount of redundant baselines that can be formed for 
the hexagonal, square and east-west layouts in the top panel are 165, 180 and 99 baselines respectively.\label{fig:geometry_function}}
\end{figure*}

@@

We have now defined enough notation to enable us to formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T. &  &\label{eq:parm_definitions}
 \end{align}
The number of model parameters that we need to solve for in Eq.~\eqref{eq:least_squares_red} is $P = 2(N+L)$, since redundant calibration is a complex problem.
We plot $L$ and $P$ for two different geometric layouts in Fig~\ref{fig:pl}.
The analytic expressions for the red and the blue curve in Fig.~\ref{fig:ln} are
\begin{equation}
L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2}, 
\end{equation}
and
\begin{equation}
L = N - 1, 
\end{equation}
respectively.

@@

\subsection{Real-and-imaginary Framework}
\label{sec:ri}
As we mentioned in Section~\ref{sec:ls} using least-squares to solve complex problems can be problematic as $\bv$ needs to be analytic in its argument. In the case 
of Eq.~\eqref{eq:least_squares_red}, \citet{Liu2010} circumvents this difficulty by 
splitting the problem into its phase-and-amplitude components.

@@

\citet{Smirnov2015} recently proposed to use the Wirtinger derivative instead of the classic real-and-imaginary approach to perform normal skymodel-based calibration. We will now show
that the Wirtinger derivative can also be applied to redundant calibration.

@@

The main flow of a single simulation is as follows. We created corrupted visibilities using the fixed simulation parameters presented in Table~\ref{tab:fixed_parm} and either one of the experimental setups summarized in Table~\ref{tab:ch_parm}. We then used per time-slot and channel redundant calibration, i.e. we use the LM algorithm, which in turn makes use of either the ADI or PCG method, to 
estimate the gains and the true observed visibilities. To obtain statistically significant results we performed, for both the ADI and PCG methods, multiple statistical independent simulations.

@@

\begin{table*}
\centering
\caption{The following fixed simulation parameters were used to generate the results of this paper.}
\begin{tabular}{|c c c c c c c|} 
\hline
Declination & Latitude & Minimum baseline & Geometry& Sources & Flux & Spatial \\
\hline 
\hline
 $-74^{\circ}39'37.481''$ & $-30^{\circ}43'17.34''$ & 20 & Hexagonal &100 & Pareto (shape param. of 2) & $U[-3,3]$ (deg.) 
\end{tabular}
\label{tab:fixed_parm}
\end{table*}

\begin{table*}
\centering
\caption{The gain error models used in this paper. We have used the symbol $x$ here as a proxy as it can either refer to time-slots or channels. We either
performed our simulations over multiple time-slots and one frequency channel or one timeslot and multiple frequencey channels (see Table~\ref{tab:ch_parm}). We use $U$ to represent 
a uniform distribution. Moreover, $c$ in the left most column denotes the speed of light.}
\begin{tabular}{|c c c c|} 
\hline
Number tag & 1 & 2 & 3\\
Model & Sinusoidal: amplitude and phase & Sinusoidal: real and imaginary parts & Linear phase slope \\ [0.5ex] 
\hline\hline
Function & $(A+1)e^{jP}$ & $A\cos(2\pi fx+B)+1.5A+jC\sin(2\pi fx+D)$ & $e^{jP}$ \\ 
\hline
Parameters & $A=a\cos(2\pi fx +b)$  & $f=5$ & $P=\tau x$ \\
 & $P =c \cos(2\pi fx +d)$ & $A,C\sim U[0.5,10]$ & $\tau = \frac{l}{c}$ \\
 & $f=5$ & $B,D\sim U[0,2\pi]$ &  $l\sim U[5,50]$ (m)\\
 & $a\sim U[0.8,0.9]$ &  & \\ 
 & $c\sim U[0.5,5]$ &  &  \\ 
 & $b,d\sim U[0,2\pi]$ &  &  \\ 
\hline
\end{tabular}
\label{tab:gain_parm}
\end{table*}

\begin{table}
\centering
\caption{We generated results using two major experimental setups. We either used one frequency channel and multiple time-slots, or one time-slot and multiple frequency 
channels. The most important parameters used in realizing these two major setups are presented here. Note that we refer to the gain error models in Table~\ref{tab:gain_parm} in the 
last row of this table.}
\begin{tabular}{|c c c|} 
\hline
 & Setup 1 & Setup 2\\
\hline
\hline
 Num. channels & 1024 & 1\\
$\nu$-range & 100-200 MHz & 1.4 GHz\\
Num. timeslots & 1 & 50\\
%Hour angle &-1^h & [-1^h,1^h]\\
Gain error model (Table~\ref{tab:gain_parm}) & 2 & 1\\
Initial guess: $\bz_0$ & $\bone$ & $\bone$\\
\hline
\end{tabular}
\label{tab:ch_parm}
\end{table}

@@

In Section~\ref{sec:red_c}, we will show that the fast redundant calibration algorithm, that is presented in \citet{Marthi2014},
is in fact nothing more than an extension of the skymodel-based ADI method, presented in \citet{Salvini2014}, to the redundant calibration use case (i.e. redundant \textsc{StEfCal}).
In doing so we will justify the naming convention we have been using up until now.

@@

\subsubsection{Skymodel-based Calibration}
\label{sec:sbc}
\textsc{StEFCal} is an ADI method in which the measurement equation, Eq.~\eqref{eq:vis_definition}, is linearized by assuming that the gains are known, but
that their conjugates are unknown \citep{Mitchell:MWA-cal,Salvini2014}. Under this assumption, the conjugates of the gains can be obtained trivially. Once the conjugates of the gains have been computed, it is possible to solve for the gains. One can then alternate between computing the gains and their conjugates until convergence
is reached. By correctly exploiting the symmetry inherent in the calibration problem, this alternating approach can be replaced by a single iterative new parameter estimation step. \cite{Salvini2014} also show that
the convergence of the algorithm can be improved, if for every even iteration, the basic new parameter estimation step is modified as follows: the current gain estimate is replaced by the average of the current gain estimate and the gain estimate of the previous odd iteration.
\textsc{StEFCal} lowers the computational complexity of normal skymodel-based calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$.

@@

\subsubsection{Redundant Calibration}
\label{sec:red_c}
In this section, we re-derive the fast redundant calibration algorithm presented in \citet{Marthi2014} using two different approaches. Both these approaches differ from the one presented in \citet{Marthi2014}. In \citet{Marthi2014}, the algorithm is derived by taking the derivative of the 
objective function $\Lambda$ relative to $\bg$ and $\by$, setting the intermediate result to naught and then solving for the unknown parameters.

@@

\subsection{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
In this section we investigate the computational complexity of the PCG method when applied to the redundant calibration problem. 
As we mentioned in Section~\ref{sec:pcg} there are two factors which influence the computational complexity of the PCG method, namely the spectral condition number and the sparsity of the matrix we wish to invert (see Eq.~\eqref{eq:cg_bound}). 
We used a $\lambda$ value of two, implying we used Eq.~\eqref{eq:LM_update} instead of Eq.~\eqref{eq:GN_update} throughout this section.
We discuss the spectral condition number and the sparsity of the modified Hessian $\bmH$, which is a non-singular matrix, in Section~\ref{sec:scn} and Section~\ref{sec:sparsity} respectively.
It is interesting to note, that the Hessian $\bH$ itself is singular which means that if we are to use Eq.~\eqref{eq:GN_update} instead we would need to be able to invert a singular matrix. In Section~\ref{sec:normal},
we showed that the PCG method would in fact be able to handle such a scenario.

@@

The evaluation of the limit in the rightmost term of Eq.~\eqref{eq:c} is evaluated as follows: 
Note that,
\begin{equation}
\label{eq:int_limit}
\lim_{N\rightarrow\infty} P^{\log_{P}(1-\gamma)} = 1-\gamma_{\infty},
\end{equation}
which is only possible if 
\begin{equation}
\lim_{N\rightarrow\infty} \log_{P}(1-\gamma) = 0,
\end{equation}
implying that $c_{\infty}$ is equal to two.

@@

\section{Redundant \textsc{StEfCal} and the ADI method}
\label{sec:red_stef_ADI}
The basic skymodel-based \textsc{StEfCal} update step is equal to  the leftmost term in Eq.~\eqref{eq:g_update} (barring $\alpha$) \citep{Salvini2014}.
Assume without any loss of generality that the array is in an east-west regular grid. Furthermore, assume that $\boldsymbol{d}$ has been re-ordered to
\begin{equation}
\widetilde{\boldsymbol{d}} = \left[d{12},\cdots,d_{N-1,N},d_{13},\cdots,d_{N-2,N},\cdots,d_{1N}\right]^T .
\end{equation}

>>>>

\begin{abstract}
In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. 
We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework.
We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton.
The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration.
It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration.
We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating 
direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. 
We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method 
and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

In this paper we propose to use the complex framework introduced by \citet{Smirnov2015} instead of the traditional amplitude-and-phase approach used by \textsc{lincal}.
We also propose to use the Levenberg-Marquardt (LM) gradient-based algorithm instead of the Gauss-Newton (GN) algorithm as it exhibits better convergence properties.
Using the LM algorithm reduces the number of algorithmic layers that are needed to perform redundant calibration.

Two ways of speeding up redundant calibration have recently been proposed. The first technique makes use of the ADI method, in which we alternate between solving the gains, their conjugates and 
the true observed visibilities \citep{Wijnholds2012,Marthi2014}. Note that the fact that \citet{Marthi2014} make use of the ADI method is not immediately clear
from the paper itself. We will elaborate on this in Section~\ref{sec:adi}. The second is to use the preconditioned conjugate gradient (PCG) method. PCG allows us to exploit the sparsity inherent in the redundant calibration problem 
to speed it up \citep{Liu2010}. A proper comparison between these two approaches has not been done before.
Therefore, we also compare the ADI \citep{Marthi2014} and PCG \citep{Liu2010} methods in this paper. We decided to focus on the complex framework proposed by \citet{Smirnov2015} rather than the traditional phase-and-amplitude approach as it 
allows us to trivially prove that the ADI method proposed by \citep{Marthi2014} is merely a special case of the more general LM algorithm.

@@

We start the paper by presenting some useful preliminaries in Section~\ref{sec:preliminaries}. In Section~\ref{sec:p_statement}, we formulate redundant calibration 
as a least squares problem, i.e. Eq~\eqref{eq:least_squares_red}. In Section~\ref{sec:main_sec} we use the complex framework proposed by \citet{Smirnov2015} to derive the GN and LM
update rules (see Eq.~\eqref{eq:GN_update} to Eq.~\eqref{eq:update}), which we can use to solve Eq~\eqref{eq:least_squares_red}. In Section~\ref{sec:results}, we compare the computational complexity of two recently proposed redundant calibration speedup algorithms with one another.
We end the paper with some final conclusions.

@@

\subsubsection{Preconditioning}
\label{sec:precon}
Preconditioning is a technique which allows us to improve the condition number of a matrix. Let $\bM$ be a positive-definite Hermitian matrix.
We can now indirectly solve for $\bx$ by using 
\begin{equation}
\label{eq:prec}
\bM^{-1}\bA\bx = \bM^{-1}\bb.
\end{equation}
If $\bM$ is a good preconditioner then $\kappa(\bM^{-1}\bA) \ll \kappa(\bA)$, i.e. a good preconditioner lowers the condition number of a matrix which in turn increases the convergence speed 
of CG (see Eq.~\eqref{eq:cg_bound}). In this paper, we denote matrix inversion with $(*)^{-1}$. The best possible choice of $\bM$ would therefore be $\bA$, which would lead to $\kappa$ becoming unity (as the largest and smallest eigenvalue of the identity matrix are equal to one).
If $\kappa$ is one then Eq.~\eqref{eq:cg_bound} simplifies and becomes $\mathbb{O}(m)$. However, if we knew the inverse of $\bA$ then obtaining $\bx$ would be trivial and there would be no need for the CG method in the first place. 
In short, a good preconditioner $\bM$ must approximate $\bA$ fairly accurately. However, if it is not much less expensive to 
compute the inverse of $\bM$ than that of $\bA$ we actually gain nothing (see Eq.~\eqref{eq:prec}).

@@

\subsection{Least Squares}
\label{sec:ls}
The method of nonlinear least squares is tantamount to solving the following optimization problem:  
\begin{equation}
\label{eq:least_squares}
\min_{\bzeta} \Lambda(\bzeta) = \min_{\bzeta} \|\brho\|_F^2 = \min_{\bzeta} \|\bdelta - \bnu(\bzeta)\|_F^2, 
\end{equation}
where $\Lambda$ is known as the objective function, $\brho$ is the residual vector, $\bdelta$ is the observed data vector, $\bnu$ is the model-predicted data vector and $\bzeta$ is the model-parameter vector that minimizes Eq.~\eqref{eq:least_squares}.
In Eq.~\eqref{eq:least_squares}, we assume that the optimization problem is real valued.
The main objective of least squares is to find the model-parameters which minimize the squared difference between the observed data and the model-predicted data.

@@

The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a standard deviation $\sigma$, where   
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}}. 
\end{equation}
Here, $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

Recall that redundant baselines sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. 
Let us now group all the baselines that measure the same $uv$-modes in a redundant array into the same redundant baseline group. Assigning the baselines of a redundant array to different redundant baseline groups has the advantage that it reduces the number of single index values we require. We now only require single index values for each redundant baseline group instead of a unique single index label for 
each baseline. Let $\phi_{pq}$ be the function that maps composite antenna pairs to the appropriate single redundant baseline group index. Note that $\phi_{pq}$ is not symmetric as 
$\phi_{pq} = 0$ if $p\geq q$. We can also define the following symmetric variant of $\phi_{pq}$:
\begin{equation}
\zeta_{pq} = 
\begin{cases}
\phi_{pq}~\textrm{if}~p \leq q\\
\phi_{qp}~\textrm{if}~p>q
\end{cases}.
\end{equation} 
It is possible to construct a simple analytic expression for $\zeta_{pq}$ when our array is in an east-west regular configuration, namely $\zeta_{pq} = |q-p|$. 
It becomes increasingly difficult to construct analytic expressions of $\zeta_{pq}$ for more complicated array layouts. 
We will refer to $\zeta_{pq}$ as the symmetric geometric function. The empirically constructed symmetric geometry
functions of three different redundant layouts are depicted in Fig.~\ref{fig:geometry_function}. We denote the range of $\zeta_{pq}$ with $\mathcal{R}(\zeta_{pq})$. The maximal element 
that $\zeta_{pq}$ can ascertain is denoted by $L$ and can be interpreted as the maximal number of unique redundant baseline groups which can be formed for a given 
array layout. Alternatively, we can interpret the symmetric geometry functions in Fig.~\ref{fig:geometry_function} as geometry matrices. In this alternative paradigm, $\zeta_{pq}$ denotes a matrix entry. The dimension of these geometry 
matrices are $N\times N$.

@@

\subfigure[Hexagonal: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./HEX_phi.pdf}\label{fig:HEX_phi}}
\subfigure[Square: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./SQR_phi.pdf}\label{fig:SQR_phi}}
\subfigure[Regular east-west: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./REG_phi.pdf}\label{fig:REG_phi}}
\caption{In the top panel we have three different redundant antenna layouts, namely a hexagonal (left), square (middle) and a regular east-west (right) layout. 
In the bottom panel we have the respective symmetric redundancy geometry functions for the layouts on the top panel. We used 91 antennas
to construct the hexagonal layout, while 100 antennas where used in the square and east-west layouts. In the case of the east-west layout we only plot the positions of the first ten antennas. The maximal amount of redundant baseline groups that can be formed for 
the hexagonal, square and east-west layouts are 165, 180 and 99 respectively.\label{fig:geometry_function}}
\end{figure*}

@@

We have now defined enough notation to enable us to formulate redundant calibration as a least-squares problem:
\begin{equation}
\label{eq:least_squares_red}
\min_{\bz} \Lambda(\bz) = \min_{\bz} \|\br\|_F^2 = \min_{\bz} \|\bd - \bv(\bz)\|_F^2, 
\end{equation}
where
\begin{align}
 \bg &=[g_1,\cdots,g_N]^T, & \by &= [y_1,\cdots,y_L]^T,\nonumber\\
 \bz &= [\bg^T,\by^T]^T. &  &\label{eq:parm_definitions}
 \end{align}
The number of model parameters that we need to solve for in Eq.~\eqref{eq:least_squares_red} is $P = 2(N+L)$, since redundant calibration is a complex problem.
We plot $L$ and $P$ for two different geometric layouts in Fig~\ref{fig:pl}.
The analytic expressions for the red and the blue curve in Fig.~\ref{fig:ln} are \citep{Camps2003}
\begin{equation}
L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2} 
\end{equation}
and
\begin{equation}
L = N - 1 
\end{equation}
respectively.

@@

\subsection{Real and imaginary Framework}
\label{sec:ri}
As we mentioned in Section~\ref{sec:ls} using least-squares to solve complex problems can be problematic as $\bv$ needs to be analytic in its argument. In the case 
of Eq.~\eqref{eq:least_squares_red}, \citet{Liu2010} circumvents this difficulty by 
splitting the problem into its phase-and-amplitude components.

@@

\citet{Smirnov2015} recently proposed to use the Wirtinger derivative instead of the classic real and imaginary approach to perform normal skymodel-based calibration. We will now show
that the Wirtinger derivative can also be applied to redundant calibration.

@@

The main flow of a single simulation is as follows. We created corrupted visibilities using the fixed simulation parameters presented in Table~\ref{tab:fixed_parm} and either one of the experimental setups summarized in Table~\ref{tab:ch_parm}. We then used per time-slot and channel redundant calibration, i.e. we use the LM algorithm, which in turn makes use of either the ADI or PCG method, to 
estimate the gains and the true observed visibilities. Our initial parameter guess $\bz_0$ was chosen to be an all one vector. To obtain statistically significant results we performed, for both the ADI and PCG methods, multiple statistical independent simulations.

@@

\begin{table*}
\centering
\caption{The following fixed simulation parameters were used to generate the results of this paper. We use $U$ to represent 
a uniform distribution.}
\begin{tabular}{|c c c c c c c|} 
\hline
Declination & Latitude & Minimum baseline & Geometry& Sources & Flux & Spatial \\
\hline 
\hline
 $-74^{\circ}39'37.481''$ & $-30^{\circ}43'17.34''$ & 20 & Hexagonal &100 & Pareto (shape param. of 2) & $U[-3,3]$ (deg.) 
\end{tabular}
\label{tab:fixed_parm}
\end{table*}

\begin{table*}
\centering
\caption{The gain error models used in this paper. We have used the symbol $x$ here as a proxy as it can either refer to time-slots or channels. We either
performed our simulations over multiple time-slots and one frequency channel or one timeslot and multiple frequencey channels (see Table~\ref{tab:ch_parm}). Moreover, $c$ in the left most column denotes the speed of light.}
\begin{tabular}{|c c c c|} 
\hline
Number tag & 1 & 2 & 3\\
Model & Sinusoidal: amplitude and phase & Sinusoidal: real and imaginary parts & Linear phase slope \\ [0.5ex] 
\hline\hline
Function & $(A+1)e^{jP}$ & $A\cos(2\pi fx+B)+1.5A+jC\sin(2\pi fx+D)$ & $e^{jP}$ \\ 
\hline
Parameters & $A=a\cos(2\pi fx +b)$  & $f=5$ & $P=\tau x$ \\
 & $P =c \cos(2\pi fx +d)$ & $A,C\sim U[0.5,10]$ & $\tau = \frac{l}{c}$ \\
 & $f=5$ & $B,D\sim U[0,2\pi]$ &  $l\sim U[5,50]$ (m)\\
 & $a\sim U[0.8,0.9]$ &  & \\ 
 & $c\sim U[0.5,5]$ &  &  \\ 
 & $b,d\sim U[0,2\pi]$ &  &  \\ 
\hline
\end{tabular}
\label{tab:gain_parm}
\end{table*}

\begin{table}
\centering
\caption{We generated results using two major experimental setups. We either used one frequency channel and multiple time-slots, or one time-slot and multiple frequency 
channels. The most important parameters used in realizing these two major setups are presented here. Note that we refer to the gain error models in Table~\ref{tab:gain_parm} in the 
last row of this table.}
\begin{tabular}{|c c c|} 
\hline
 & Setup 1 & Setup 2\\
\hline
\hline
 Num. channels & 1024 & 1\\
$\nu$-range & 100-200 MHz & 1.4 GHz\\
Num. timeslots & 1 & 50\\
%Hour angle &-1^h & [-1^h,1^h]\\
Gain error model (Table~\ref{tab:gain_parm}) & 2 & 1\\
%Initial guess: $\bz_0$ & $\bone$ & $\bone$\\
\hline
\end{tabular}
\label{tab:ch_parm}
\end{table}

@@

In Section~\ref{sec:red_c}, we will show that the fast redundant calibration algorithm, that is presented in \citet{Marthi2014},
is in fact nothing more than an extension of the skymodel-based ADI method, presented in \citet{Salvini2014}, to the redundant calibration use case (i.e. redundant \textsc{StEfCal}).
In doing so we will justify the naming convention we have been using for it.

@@

\subsubsection{Skymodel-based Calibration}
\label{sec:sbc}
\textsc{StEfCal} is an ADI method in which the measurement equation, Eq.~\eqref{eq:vis_definition}, is linearized by assuming that the gains are known, but
that their conjugates are unknown \citep{Mitchell:MWA-cal,Salvini2014}. Under this assumption, the conjugates of the gains can be obtained trivially. Once the conjugates of the gains have been computed, it is possible to solve for the gains. One can then alternate between computing the gains and their conjugates until convergence
is reached. By correctly exploiting the symmetry inherent in the calibration problem, this alternating approach can be replaced by a single iterative new parameter estimation step. \cite{Salvini2014} also show that
the convergence of the algorithm can be improved, if for every even iteration, the basic new parameter estimation step is modified as follows: the current gain estimate is replaced by the average of the current gain estimate and the gain estimate of the previous odd iteration.
\textsc{StEFCal} lowers the computational complexity of normal skymodel-based calibration from $\mathbb{O}(N^3)$ to $\mathbb{O}(N^2)$.

@@

\subsubsection{Redundant Calibration}
\label{sec:red_c}
In this section, we re-derive the fast redundant calibration algorithm presented in \citet{Marthi2014} using two different approaches. Both these approaches differ from the one presented in \citet{Marthi2014}. In \citet{Marthi2014}, the algorithm is derived by taking the derivative of the 
objective function $\Lambda$ relative to $\bg$ and $\by$, setting the intermediate results to naught and then solving for the unknown parameters.

@@

\subsection{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
In this section we investigate the computational complexity of the PCG method when applied to the redundant calibration problem. 
As we mentioned in Section~\ref{sec:conj_grad} there are two factors which influence the computational complexity of the PCG method, namely the spectral condition number and the sparsity of the matrix we wish to invert (see Eq.~\eqref{eq:cg_bound}). 
We used a $\lambda$ value of two, implying we used Eq.~\eqref{eq:LM_update} instead of Eq.~\eqref{eq:GN_update} throughout this section.
We discuss the spectral condition number and the sparsity of the modified Hessian $\bmH$, which is a non-singular matrix, in Section~\ref{sec:scn} and Section~\ref{sec:sparsity} respectively.
It is interesting to note, that the Hessian $\bH$ itself is singular which means that if we are to use Eq.~\eqref{eq:GN_update} instead we would need to be able to invert a singular matrix. In Section~\ref{sec:normal},
we showed that the PCG method would in fact be able to handle such a scenario.

@@

The evaluation of the limit in the rightmost equation of Eq.~\eqref{eq:c} is evaluated as follows: 
note that
\begin{equation}
\label{eq:int_limit}
\lim_{N\rightarrow\infty} P^{\log_{P}(1-\gamma)} = 1-\gamma_{\infty},
\end{equation}
which is only possible if 
\begin{equation}
\lim_{N\rightarrow\infty} \log_{P}(1-\gamma) = 0,
\end{equation}
implying that $c_{\infty}$ is equal to two.

@@

\section{Redundant \textsc{StEfCal} and the ADI method}
\label{sec:red_stef_ADI}
The basic skymodel-based \textsc{StEfCal} update step is equal to  the leftmost term in Eq.~\eqref{eq:g_update} (barring $\alpha$) \citep{Salvini2014}.
Assume without any loss of generality that the array is in an east-west regular grid. Furthermore, assume that $\boldsymbol{d}$ (see Eq.~\eqref{eq:vis_red_vec}) has been re-ordered to
\begin{equation}
\widetilde{\boldsymbol{d}} = \left[d{12},\cdots,d_{N-1,N},d_{13},\cdots,d_{N-2,N},\cdots,d_{1N}\right]^T .
\end{equation}
The vector $\widetilde{\bn}$ should be interpreted in a similar manner.
